Quantum Hoare Type Theory

### 2020-10-26 ms-paper/ms-paper.tex
emp -> \top

\lstset{
	basicstyle=\small\ttfamily,
	flexiblecolumns=false,
	basewidth={0.5em,0.45em},
	commentstyle=\color{dkgreen},
	keywordstyle=\bfseries\color{blue},
	stringstyle=\color{violet},
	literate={+}{{$+$}}1 {/}{{$/$}}1 {*}{{$*$}}1 {=}{{$=$}}1
	{>}{{$>$}}1 {<}{{$<$}}1 {\\}{{$\lambda$}}1
	{\\\\}{{\char`\\\char`\\}}1
	{->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
	{<=}{{$\Leftarrow$}}2 {=>}{{$\Rightarrow$}}2
	{\ .}{{$\circ$}}2 {\ .\ }{{$\circ$}}2
	{>>}{{>>}}2 {>>=}{{>>=}}2
	{|}{{$\vert$}}1 {\\>}{{$\rangle$}}1
	{/\\}{{$\wedge$}}1 {\\/}{{$\vee$}}1
	{\\in}{{$\in$}}1 {\\o}{{$\circ$}}1 {\\Pi}{{$\Pi$}}1
	{\\Phi+}{{$\Phi^+$}}2
	{|->}{{$\mapsto$}}2 {-o}{{$\multimap$}}2
	{o*}{{$\otimes$}}1
}

@@

\begin{mathpar}
	\inferrule
	{\Delta \vdash N \chk \type{Qbit}
		\\ \Delta; \Psi; P \prove (N \hookrightarrow -)
		\\ \Delta, x:\type{Bool}; P \circ ((N \mapsto -) \multimap \kw{emp}) \vdash E \syn y:B . Q}
	{\Delta; P \vdash x \Leftarrow \kw{measure} N; E \syn y:B. (\exists x:\type{Bool}.Q)}
\end{mathpar}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
hqw : {emp} r : Bool {emp /\ Id(r, false)}
    = do q <= init false;
         return measure q
\end{lstlisting}
\end{minipage}

In this trivial program where we initialize a new qubit with \texttt{false} and then immediately measure it, we assert that the result is equal to \texttt{false}. The complete specification also implicitly says that the existing quantum state is not affected in any way as both the pre and postcondition include the \texttt{emp} assertion.

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
hqw : {emp} r : Bool {emp /\ Id(r, false)}
-- P0: emp
    = do q <= init false;
-- P1: P0 \o (q |-> |0\>)
         return measure q
-- P2: P1 \o ((q |-> -) -o emp)
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rnd : {emp} r : Bool {emp}
    = do q <= init false;
         apply H to q;
         return measure q
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rnd : {emp} r : Bool {emp}
-- P0: emp
    = do q <= init false;
-- P1: P0 \o (q |-> |0\>)
         apply H to q;
-- P2: P1 \o ((q |-> |0\>) -o (q |-> |+\>))
         return measure q
-- P3: P2 \o ((q |-> -) -o emp)
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}  -- a,b \in {true, false}
         = do qa <= init false;
              apply H to qa;
              qb <= init false;
              apply CNOT to [qa, qb];
              return (measure qa, measure qb)
\end{lstlisting}
\end{minipage}

@@

\begin{lstlisting}[language=QHaskell]
testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}
              -- P0: emp
         = do qa <= init false;
              -- P1: P0 \o (qa |-> |0\>)
              apply H to qa;
              -- P2: P1 \o ((qa |-> |0\>) -o (qa |-> |+\>))
              qb <= init false;
              -- P3: P2 \o (qb |-> |0\>)
              apply CNOT to [qa, qb];
              -- P4: P3 \o ((qa |-> |+\>, qb |-> |0\>) -o (qa, qb) |-> |\Phi+\>)
              return (measure qa, measure qb)
              -- P5: P4 \o ((qa |-> -) -o emp) \o ((qb |-> -) -o emp)
\end{lstlisting}

@@

\begin{turn}{90}
	\begin{minipage}{42pc}
	\begin{mathpar}
		\inferrule*[right=HIntro]
		{\inferrule*[right=consq]
			{\inferrule*[right=init]
				{\inferrule*[right=measure]
					{\inferrule*[right=return]
						{q : \type{Qbit}, r : \type{Bool} \vdash r \chk \type{Bool}}
						{q : \type{Qbit}, r : \type{Bool}; P_1 \circ ((q \mapsto -) \multimap emp) \vdash \kw{return} r \syn r : \type{Bool} . R}
						\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
						\\ \cdot; P_0 * (q \mapsto \ket{0}) \prove (q \hookrightarrow -)
					}
					{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measure} q; E_3 \syn r : \type{Bool} . (\exists r : \type{Bool}.R)}
					\\ \cdot \vdash \kw{false} \chk \type{Bool}
				}
				{
					\cdot; P_0 \vdash q \Leftarrow \kw{init false}; E_{23} \syn r : \type{Bool} . (\exists q : \type{Qbit}.R)
				}
				\\ r : \type{Bool}; i,m ; R \prove emp \multimap emp \wedge Id(r, \kw{false})
			}
			{
				\cdot; \lambda i . \lambda m . i = m \wedge (emp * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bool} . (emp \multimap emp \wedge Id(r, \kw{false}))
			}
		}
		{\cdot \vdash \kw{do} E \chk \{emp\} r : \type{Bool} \{emp \wedge \kw{Id}(r, \kw{false})\}}
	\end{mathpar}
	\end{minipage}
\end{turn}

@@

\subsection{Hadamard basis states}
\begin{lstlisting}[language=QHaskell]
qplus : {emp} r : Qbit {Id(r, |+\>)}
      = do q <= init false;
           apply H to q;
           return q

qminus : {emp} r : Qbit {Id(r, |-\>)}
       = do q <= init true;
            apply H to q;
            return q
\end{lstlisting}

@@

\subsection{Bell pair}
\begin{lstlisting}[language=QHaskell]
bell : {emp} (a, b) : (Qbit o* Qbit) {Id(a, b) /\ a \in {|0\>, |1\>}}
     = do qa <- qplus;
          qb <- share qa;
          return (qa, qb)
\end{lstlisting}

\subsection{Testing modular Bell pair}
\begin{lstlisting}[language=QHaskell]
testBell : {emp} (a, b) : (Bool o* Bool) {emp /\ Id(a, b)}
         = do (qa, qb) <- bell;
              return (measure qa, measure qb)
\end{lstlisting}

\section{Teleportation}
\label{app:teleport}
\todo{TODO: Refine assertions}
\subsection{Alice's circuit}
\begin{lstlisting}[language=QHaskell]
alice : \Pi a : Qbit. \Pi e : Qbit
        {(Id(a, -) /\ entangled(e)}
           r : (Bool o* Bool)
        {emp}
      = \a.\e.do apply CNOT to [a, e];
                 apply H to a;
                 return (measure a, measure e)
\end{lstlisting}

>>>>

\lstset{
	basicstyle=\small\ttfamily,
	flexiblecolumns=false,
	basewidth={0.5em,0.45em},
	commentstyle=\color{dkgreen},
	keywordstyle=\bfseries\color{blue},
	stringstyle=\color{violet},
	literate={+}{{$+$}}1 {/}{{$/$}}1 {*}{{$*$}}1 {=}{{$=$}}1
	{>}{{$>$}}1 {<}{{$<$}}1 {\\}{{$\lambda$}}1
	{\\\\}{{\char`\\\char`\\}}1
	{->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
	{<=}{{$\Leftarrow$}}2 {=>}{{$\Rightarrow$}}2
	{\ .}{{$\circ$}}2 {\ .\ }{{$\circ$}}2
	{>>}{{>>}}2 {>>=}{{>>=}}2
	{|}{{$\vert$}}1 {\\>}{{$\rangle$}}1
	{/\\}{{$\wedge$}}1 {\\/}{{$\vee$}}1
	{\\in}{{$\in$}}1 {\\o}{{$\circ$}}1 {\\Pi}{{$\Pi$}}1
	{\\Phi+}{{$\Phi^+$}}2
	{|->}{{$\mapsto$}}2 {-o}{{$\multimap$}}2
	{o*}{{$\otimes$}}1
	{\\top}{{$\top$}}1
	{\\bot}{{$\bot$}}1
}

@@

\begin{mathpar}
	\inferrule
	{\Delta \vdash N \chk \type{Qbit}
		\\ \Delta; \Psi; P \prove (N \hookrightarrow -)
		\\ \Delta, x:\type{Bool}; P \circ ((N \mapsto -) \multimap \top) \vdash E \syn y:B . Q}
	{\Delta; P \vdash x \Leftarrow \kw{measure} N; E \syn y:B. (\exists x:\type{Bool}.Q)}
\end{mathpar}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
hqw : {\top} r : Bool {\top /\ Id(r, false)}
    = do q <= init false;
         return measure q
\end{lstlisting}
\end{minipage}

In this trivial program where we initialize a new qubit with \texttt{false} and then immediately measure it, we assert that the result is equal to \texttt{false}. The complete specification also implicitly says that the existing quantum state is not affected in any way as both the pre and postcondition include the $\top$ assertion.

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
hqw : {\top} r : Bool {\top /\ Id(r, false)}
-- P0: \top
    = do q <= init false;
-- P1: P0 \o (q |-> |0\>)
         return measure q
-- P2: P1 \o ((q |-> -) -o \top)
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rnd : {\top} r : Bool {\top}
    = do q <= init false;
         apply H to q;
         return measure q
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rnd : {\top} r : Bool {\top}
-- P0: \top
    = do q <= init false;
-- P1: P0 \o (q |-> |0\>)
         apply H to q;
-- P2: P1 \o ((q |-> |0\>) -o (q |-> |+\>))
         return measure q
-- P3: P2 \o ((q |-> -) -o \top)
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
testBell : {\top} (a, b) : (Bool, Bool) {\top /\ Id(a, b)}  -- a,b \in {true, false}
         = do qa <= init false;
              apply H to qa;
              qb <= init false;
              apply CNOT to [qa, qb];
              return (measure qa, measure qb)
\end{lstlisting}
\end{minipage}

@@

\begin{lstlisting}[language=QHaskell]
testBell : {\top} (a, b) : (Bool, Bool) {\top /\ Id(a, b)}
              -- P0: \top
         = do qa <= init false;
              -- P1: P0 \o (qa |-> |0\>)
              apply H to qa;
              -- P2: P1 \o ((qa |-> |0\>) -o (qa |-> |+\>))
              qb <= init false;
              -- P3: P2 \o (qb |-> |0\>)
              apply CNOT to [qa, qb];
              -- P4: P3 \o ((qa |-> |+\>, qb |-> |0\>) -o (qa, qb) |-> |\Phi+\>)
              return (measure qa, measure qb)
              -- P5: P4 \o ((qa |-> -) -o \top) \o ((qb |-> -) -o \top)
\end{lstlisting}

@@

\begin{turn}{90}
	\begin{minipage}{42pc}
	\begin{mathpar}
		\inferrule*[right=HIntro]
		{\inferrule*[right=consq]
			{\inferrule*[right=init]
				{\inferrule*[right=measure]
					{\inferrule*[right=return]
						{q : \type{Qbit}, r : \type{Bool} \vdash r \chk \type{Bool}}
						{q : \type{Qbit}, r : \type{Bool}; P_1 \circ ((q \mapsto -) \multimap \top) \vdash \kw{return} r \syn r : \type{Bool} . R}
						\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
						\\ \cdot; P_0 * (q \mapsto \ket{0}) \prove (q \hookrightarrow -)
					}
					{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measure} q; E_3 \syn r : \type{Bool} . (\exists r : \type{Bool}.R)}
					\\ \cdot \vdash \kw{false} \chk \type{Bool}
				}
				{
					\cdot; P_0 \vdash q \Leftarrow \kw{init false}; E_{23} \syn r : \type{Bool} . (\exists q : \type{Qbit}.R)
				}
				\\ r : \type{Bool}; i,m ; R \prove \top \multimap \top \wedge Id(r, \kw{false})
			}
			{
				\cdot; \lambda i . \lambda m . i = m \wedge (\top * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bool} . (\top \multimap \top \wedge Id(r, \kw{false}))
			}
		}
		{\cdot \vdash \kw{do} E \chk \{\top\} r : \type{Bool} \{\top \wedge \kw{Id}(r, \kw{false})\}}
	\end{mathpar}
	\end{minipage}
\end{turn}

@@

\subsection{Hadamard basis states}
\begin{lstlisting}[language=QHaskell]
qplus : {\top} r : Qbit {Id(r, |+\>)}
      = do q <= init false;
           apply H to q;
           return q

qminus : {\top} r : Qbit {Id(r, |-\>)}
       = do q <= init true;
            apply H to q;
            return q
\end{lstlisting}

@@

\subsection{Bell pair}
\begin{lstlisting}[language=QHaskell]
bell : {\top} (a, b) : (Qbit o* Qbit) {Id(a, b) /\ a \in {|0\>, |1\>}}
     = do qa <- qplus;
          qb <- share qa;
          return (qa, qb)
\end{lstlisting}

\subsection{Testing modular Bell pair}
\begin{lstlisting}[language=QHaskell]
testBell : {\top} (a, b) : (Bool o* Bool) {\top /\ Id(a, b)}
         = do (qa, qb) <- bell;
              return (measure qa, measure qb)
\end{lstlisting}

\section{Teleportation}
\label{app:teleport}
\todo{TODO: Refine assertions}
\subsection{Alice's circuit}
\begin{lstlisting}[language=QHaskell]
alice : \Pi a : Qbit. \Pi e : Qbit
        {(Id(a, -) /\ entangled(e)}
           r : (Bool o* Bool)
        {\top}
      = \a.\e.do apply CNOT to [a, e];
                 apply H to a;
                 return (measure a, measure e)
\end{lstlisting}
